Calibrating seed-based heuristics to map short DNA reads

### 2019-04-25 manuscript.tex
Added edits from eduardvalera@gmail.com. Vesion submitted to bioRxiv.

\title{On mapping errors and seeding heuristics for
DNA alignment}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze the data. The fate of most reads is to be mapped to
a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. In the seeding
approach, short matches between the reads and the genome are used to
narrow the search to a set of candidate locations. Several seeding
variants used in modern mappers show good empirical performance but they
are difficult to calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that reads are
mismapped due to failures at the seeding step. We describe the properties
of simple exact seeds, skip-seeds and MEM seeds (Maximal Exact
Match).
% Innovation.
The major innovation is to use concepts from analytic combinatorics to
represent reads as abstract sequences, and to specify their generative
function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a practicial
solution to the problem of calibrating seeding heuristic for short reads.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
\end{abstract}

@@

Because sequencing errors can convert the sequence of the target into one
of its duplicates, the \emph{true} location of a DNA fragment is not
always the \emph{best} (as measured by the identity between the sequence
and the candidate location). This leads to the following question: how can
we identify the \emph{best} location of the fragment in the genome?

@@

There exists algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run fast, but that may
return an incorrect result~\cite{Waterman1984}.

@@

If seeding is fast, we can quickly zoom in on a small set of candidate
locations and reduce the input size of the (slow) alignment algorithm. The
disadvantage is that the target location is not always in the candidate
set. When that happens, the read cannot be mapped to the correct location
because it has been filtered out at the seeding step.

The slowness of exact alignment algorithms imposes a trade-off between
speed and accuracy: If the seeding step is set to produce many candidates,
it is likely to retrieve the candidate. The downside is that this will
take long. Conversely, if the seeding step is set to produce few
candidates, the process will run faster but the target is less likely to
be in the candidate set.

@@

Before going further, we introduce some vocabulary that will be useful to
simplify the exposition of the theory.

@@

This observation is fundamental: in order to estimate the error rate of of
a mapping algorithm, we must know how often the associated seeding
heuristic is off target. For lack of theoretical framework, this
information is missing, meaning that current mappers cannot be properly
calibrated.

Our focus here is to develop a theory to estimate the probability that
seeding is off target, when using the most popular seeding heuristics for
short reads. Previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to control mapping errors. Other authors investigated
the reliability of mapping algorithms~\cite{pmid23872968}, but they
focused on the probability that random sequences have a single hit,
recognizing that solving the problem requires taking into consideration
the repeat structure of the genome. Here we address this problem in
general terms and we provide a practical way to estimate the mismapping
rate in complex genomes.

The rest of this document is organized to present the content in a
didactic manner. The order of the examples is meant to facilitate the
aquisition of the main concepts, meaning that we sometimes go back and
forth between different types of seeding problems. We first give some
background on the different types of seeding methods considered here, we
then introduce the computational strategy based on transfer matrices, and
then delve into each specific seeding method.

@@

The term ``seed'' changes meaning with different authors. It can designate
a part of the read, a part of the genome, a particular sequence motif or a
structured pattern of matches. Also, the term does not always imply an
exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
uses ``spaced seeds'' that tolerate mismatches. To avoid potential
confusion, we will adopt the following convention.

\begin{definition}
A seed is a subsequence of the read that is above a certain size $\gamma$
(defined by the context of the problem) and that has at least one perfect
match in the reference genome.
\end{definition}

When a seed matches a particular location of the genome, we will say that
it is a ``seed for that location''. For instance, we will often use
expressions such as ``a seed for the target'' or ``a seed for a
duplicate''.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately there
exists practical methods to solve it, even when the reference genome is
very large.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds described below, it is
even the only practical option~\cite{pmid24336412, pmid25399029,
pmid23349213, pmid19389736}. However, the detail if of little interest
here: we simply assume that seeds are known at all times without ambiguity
because this problem has several practical solutions.

@@

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
overlapping matches in the same regions, as we are about to illustrate.

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three duplicates and the read has three miscalled
nucleotides (indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above the actual molecule
(Target), and above homologous sequences also present in the genome
(Duplicates). Sequencing errors are indicated by a star, each marked
\texttt{T} was actually an \texttt{A} that was misread. The mismatches
between the genomic sequences and the read are indicated in bold. The
exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
above the read. The matching regions in the genomic sequences are
shadowed in grey. Several overlapping seeds accumulate at the center of
the read, which is typical for this seeding method.}
\label{fig:exact_seed_example}
\end{figure}

The exact seeds with $\gamma=7$ are shown above the sequence of the read.
Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to an off-target seed.
However, erroneous nucleotides that are mismatches for \emph{all} the
sequences cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of seeds.

The middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive compared to spaced seed~\cite{pmid11934743} so it
is used only in few specific applications. Nevertheless, it will be
extremely useful for the development of the present theory.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
missing seeds are represented by dotted lines. Skipping nucleotides
reduces the clutter at the center of the read, but it also removes the
only seed for the second duplicate.}
\label{fig:skip_seed_example}
\end{figure}

@@

MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
to define. Unlike exact seeds and skip seeds, their size is variable. They
are used in BWA-MEM~\cite{li2013aligning} where they give good empirical
results. To describe MEM seeds, let us first focus on MEMs.

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location in this case (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
at the bottom effectively hide the target, so it cannot be discovered.
This can occur even when the true location is the best and when there is a
single error on the read.}
\label{fig:full_masking_example}
\end{figure}

@@

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaneously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each nucleotide
position, any given duplicate is identical to the target with probability
$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
one the remaining three nucleotides (\textit{i.e.} each is found with
probability $\mu/3$).

@@

As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
we recode the reads in alphabets of custom symbols and we specify a
construction plan of the reads using a so-called \emph{transfer matrix}
$M(z)$. The transfer matrix specifies which types of segments can follow
each other in the reads of interest: the entry at coordinates $(i,j)$ is
the weighted generating function of segments of type $j$ that can be
appended to segments of type $i$.

From the basic operations on weighted generating functions, the entry of
$M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$ at
coordinates $(i,j)$ is the weighted generating function of all the reads
that end with a segment of type $j$ and that can be appended to a segment
of type $i$. The examples below will clarify the key steps of this
strategy.

@@

This recoding allows us to partition the read in am important way.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
An example read is represented in the mismatch alphabet. The symbol
$\Downarrow$ represents a mismatch against the target (it is an erroneous
nucleotide) and the symbol $\square$ represents a match (it is a correct
nucleotide). The symbol $|$ is appended to the end of the read. The
symbolic sequence of the target is represented below, where an open square
stands for a match and a closed square stands for a mismatch.}
\label{fig:simple}
\end{figure}

@@

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term --- associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by $\Downarrow$ segments and only
by those (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. It is easy to check that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} = a_0 + a_1z + a_2z^2 +
\ldots
\end{equation}

The coefficients of the Taylor series $a_0 + a_1z + a_2z^2 + \ldots$ are
unknown. But by construction, $a_k$ is the quantity of interest,
\textit{i.e.} it is the probability that a read of size $k$ does not
contain an on-target seed, so we now need to extract the coefficients
from expression (\ref{eq:F}). There are several methods to do so; the one
we choose here is to build a recurrence equation to compute $a_k$ from the
previous terms of the series $a_0, a_1, \ldots, a_{k-1}$. For this, we
observe that for $z \neq 1/q$, equation (\ref{eq:F}) is equivalent to
\begin{equation*}
(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
\end{equation*}
Balancing the terms of same degree on both sides of the equation leads
to the following relationship between the coefficients of the series:
\begin{equation}
\label{eq:recur}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -pq^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}

@@

As in the previous example, we need to know whether a nucleotide is a
sequencing error, but this time we also need to know its phase in the
repeated cycles of skipped positions. For this, we define the skip-$n$
alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
\ldots, \Downarrow_n \}$. The symbol $\square$ still represents a correct
nucleotide and the symbol $|$ is still a terminator added at the end of
the read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent
sequencing errors and $i$ indicates the number of nucleotides till the
next non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
before a non-skipped position and $i=n$ for nucleotides at a non-skipped
position).

@@

As mentioned above, after the terminator $\Downarrow_i$, we can append
segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
those $\gamma+i$ possible segments are distributed according to the laws
of the arithmetic modulo $n+1$. For $i > 0$ for instance, if the next
segment has no symbol $\square$ then it is terminated by
$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
involves a substantial amount of bookkeeping. One can check that the
expression of the transfer matrix $M_n(z)$ is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{eqnarray}
H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
\end{eqnarray}
where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.

@@

Since the duplicates are assumed to evolve independently of each other and
through the same mutagenesis process, the events are conditionally
independent (we do not have full independence because $S_0$ depends on
sequencing errors only whereas the events $S_j$ depend on sequencing
errors and on the divergence between duplicates). We can thus write
\begin{eqnarray}
\notag
P(\overline{S}_0 \cap \ldots \cap \overline{S}_N) &=&
  P(\overline{S}_0) \cdot P(\overline{S}_1 \cap \ldots \cap
  \overline{S}_N | \overline{S}_0) \\
\label{eq:Pnull}
  &=& P(\overline{S}_0) \cdot P(\overline{S}_1 | \overline{S}_0)^N =
  P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0 \cap
\overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{eqnarray}
Combinging the two equations above, we obtain
\begin{equation}
\label{eq:Poff}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{equation}

The probability that seeding is off target thus depends on only two
quantitites. We have already seen how to compute $P(\overline{S}_0)$ in
section~\ref{sec:example_exact} using recursive expression
(\ref{eq:recur}). We now need to find a way to compute $P(\overline{S}_0
\cap \overline{S}_1)$.

@@

\begin{definition}
At a given position of the read, a duplicate is a \emph{hard mask} if its
match length is strictly longer than the match length of the target. A
duplicate is a \emph{soft mask} if it has the same match length as the
target.
\end{definition}

>>>>

\title{Calibrating seed-based heuristics to map short DNA reads}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies creates a need
for faster algorithms. The fate of most reads is to be mapped to a
reference sequence, typically a genome. Modern mappers rely on heuristics
to gain speed at a reasonable cost for accuracy. In the seeding approach,
short matches between the reads and the genome are used to narrow the
search to a set of candidate locations. Several seeding variants used in
modern mappers show good empirical performance but they are difficult to
calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that reads are
mapped to a wrong location due to limitations at the seeding step. We
describe the properties of simple exact seeds, skip-seeds and MEM seeds
(Maximal Exact Match).
% Innovation.
The main innovation of this work is to use concepts from analytic
combinatorics to represent reads as abstract sequences, and to specify
their generative function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a practical
solution to the problem of calibrating seeding heuristics for short reads.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
\end{abstract}

@@

Because sequencing errors can convert the sequence of the target into one
of its duplicates, the \emph{true} location of a DNA fragment is not
always the \emph{best} candidate (as measured by the identity between the
sequence and the candidate location). This leads to the following
question: how can we identify the \emph{best} location of the fragment in
the genome?

@@

There exist algorithms to solve the best location
problem~\cite{pmid7265238,pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.} algorithms that run faster, but that may
return an incorrect result~\cite{Waterman1984}.

@@

If seeding is fast, we can quickly zoom in on a small set of candidate
locations and reduce the input size of the (slow) alignment algorithm. The
disadvantage is that the target location is not always detected and
included in the candidate set. When that happens, the read cannot be
mapped to the true location because it has been filtered out at the
seeding step.

The slowness of exact alignment algorithms imposes a trade-off between
speed and accuracy: If the seeding step is set to produce many candidates,
it is likely to retrieve the true location. The downside is that
processing all the candidates would take long. Conversely, if the seeding
step is set to produce few candidates, the process would run faster but
the target is less likely to be in the candidate set.

@@

Before going further, we introduce some nomenclature that will be useful
to simplify the exposition of the theory.

@@

This observation is fundamental: in order to estimate the error rate of a
mapping algorithm, we must know how often the associated seeding heuristic
is off target. For lack of theoretical framework, this information is
missing, meaning that current mappers are not properly calibrated.

Our focus here is to develop a theory to estimate the probability that
seeding is off target, when using the most popular seeding heuristics for
short reads. Previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to estimate the frequency of mapping errors. Other
authors investigated the reliability of mapping
algorithms~\cite{pmid23872968}, but they focused on the probability that
random sequences have a single hit, recognizing that solving the problem
requires taking into consideration the repeat structure of the genome.
Here we address this problem in general terms and we provide a practical
way to estimate the mismapping rate in complex genomes.

The rest of this document is organized to present the content in a
didactic manner. The order of the examples is meant to facilitate the
acquisition of the main concepts, meaning that we sometimes go back and
forth between different types of seeding strategies. We first give some
background on the different types of seeding methods considered here; we
then introduce the computational strategy based on transfer matrices;
finally we delve into each specific seeding method.

@@

The term ``seed'' changes meaning with different authors. It can designate
a part of the read, a part of the genome, a particular sequence motif or a
structured pattern of matches. Also, the term does not always imply an
exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
uses ``spaced seeds'' that tolerate mismatches. To avoid potential
confusion, we will adopt the following convention:

\begin{definition}
A seed is a subsequence of the read that has size at least $\gamma$
(defined by the context of the problem) and that has at least one perfect
match in the reference genome.
\end{definition}

When a seed matches a particular location of the genome, we say that it is
a ``seed for that location''. For instance, we often use expressions such
as ``a seed for the target'' or ``a seed for a duplicate''.

This definition presents a computational challenge: to know if a given
subsequence is a seed we need to know if it exists somewhere in the
genome. This is a non trivial problem in itself, but fortunately we can
use practical methods to solve it, even when the reference genome is very
large.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of fixed size $k$
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, some text-indexing structures have no set size
so they can be used to search seeds of different lengths. For instance,
the FM-index is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint~\cite{ferragina2000opportunistic,
ferragina2005indexing}.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds, it is even the only
practical option~\cite{pmid24336412, pmid25399029, pmid23349213,
pmid19389736}. However, the detail is of little interest here: we simply
assume that seeds are known at all times without ambiguity because this
problem has several practical solutions.

@@

Exact seeds have a fixed size $\gamma$. When using exact seeds, all the
perfect matches of size $\gamma$ are collected to constitute the candidate
set. This seeding heuristic was used in the first version of
BLASTN~\cite{pmid2231712}, but is no longer used because it produces many
overlapping matches within the same regions.

Fig.~\ref{fig:exact_seed_example} shows a read produced by an imperfect
sequencer. The sequenced DNA fragment comes from a region of the genome
that has three duplicates and the read has three miscalled nucleotides
(indicated by stars).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{exact_seed_example.pdf}
\caption{\textbf{Example of exact seeds.}
A sequenced DNA fragment (Read) is shown above the actual molecule
(Target), and above homologous sequences also present in the genome
(Duplicates). Sequencing errors are indicated by a star (each marked
\texttt{T} was actually an \texttt{A} that was misread). The mismatches
between the genomic sequences and the read are indicated in bold. The
exact seeds of size $\gamma = 7$ are indicated as horizontal grey lines
above the read. The matching regions in the genomic sequences are
shadowed in grey. Several overlapping seeds accumulate at the center of
the read, which is typical for this seeding method.}
\label{fig:exact_seed_example}
\end{figure}

Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to an off-target seed.
However, sequencing errors that are mismatches for \emph{all} the
sequences cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of seeds.

The middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive compared to spaced seeds~\cite{pmid11934743} so it
is used only in few specific applications. Nevertheless, it will be
extremely useful for the development of the present theory.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=1]{skip_seed_example.pdf}
\caption{\textbf{Example of skip seeds.}
The sequences and the annotations are the same as in
Fig.~\ref{fig:exact_seed_example}, but here we use skip-1 seeds. In other
words, seeds can never start at nucleotides 2, 4, 6 \textit{etc}. To
highlight the difference with Fig.~\ref{fig:exact_seed_example}, the
missing seeds are represented by dotted lines.}
\label{fig:skip_seed_example}
\end{figure}

@@

MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
to define. Unlike exact seeds and skip seeds, their size is variable. They
are used in BWA-MEM~\cite{li2013aligning} where they give good empirical
results. To describe MEM seeds, let us first clarify the meaning of
``Maximal Exact Match''.

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location using MEM seeds (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
at the bottom effectively hide the target, so it cannot be discovered.
This can occur even when the true location is the best candidate and when
there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

@@

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaneously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events. Instead, we simply assume that at each nucleotide
position, any given duplicate is identical to the target with probability
$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
one the three remaining nucleotides (\textit{i.e.} each is found with
probability $\mu/3$).

@@

As explained in~\cite{filion2017analytic} and~\cite{filion2018analytic},
we recode the reads in alphabets of custom symbols and we specify a
construction plan of the reads using a \emph{transfer matrix} $M(z)$. The
transfer matrix specifies which types of segments can follow each other in
the reads of interest: the entry at coordinates $(i,j)$ is the weighted
generating function of segments of type $j$ that can be appended to
segments of type $i$.

$M(z)$ contains the weighted generating functions of all the reads that
consist of a single fragment. From the basic operations on weighted
generating functions, $M(z)^s$ contains the weighted generating functions
of all the reads that consist of $s$ segments. Therefore, the entry at
coordinates $(i,j)$ of the matrix $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z)
\cdot (I-M(z))^{-1}$ is the weighted generating function of the reads of
any size and any number of segments, that end with a segment of type $j$
and that can be appended to a segment of type $i$. The examples below will
clarify the key steps of this strategy.

@@

This recoding allows us to partition the read in an important way.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
An example read is represented in the mismatch alphabet. The symbol
$\Downarrow$ represents a mismatch against the target (an erroneous
nucleotide) and the symbol $\square$ represents a match (a correct
nucleotide). The symbol $|$ is appended to the end of the read. The
symbolic sequence of the target is represented below, where an open square
stands for a match and a closed square stands for a mismatch.}
\label{fig:simple}
\end{figure}

@@

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term --- associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by $\Downarrow$ segments and only
by those (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. It is easy to check that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)}.
\end{equation}

This function can be expanded as a Taylor series $a_0 + a_1z + a_2z^2 +
\ldots$, but the coefficients $a_0, a_1, a_2, \ldots$ are unknown. By
construction, $a_k$ is the quantity of interest, \textit{i.e.} it is the
probability that a read of size $k$ does not contain an on-target seed, so
we now need to extract the coefficients from expression (\ref{eq:F}).
There are several methods to do so; the one we choose here is to build a
recurrence equation to compute $a_k$ from the previous terms of the series
$a_0, a_1, \ldots, a_{k-1}$. For this, we observe that for $z \neq 1/q$,
equation
\begin{equation*}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} =
  a_0 + a_1z +a_2z^2 + \ldots
\end{equation*}
is equivalent to
\begin{equation*}
(1-qz) \left(1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)\right)
(a_0 + a_1z +a_2z^2 + \ldots) = 1-(qz)^\gamma.
\end{equation*}
Balancing the terms of same degree on both sides of the equation leads
to the following relationship between the coefficients of the series:
\begin{equation}
\label{eq:recur}
a_k = 
\begin{cases}
1            &\quad\text{if } k < \gamma, \\
1 -q^\gamma &\quad\text{if } k = \gamma, \\
a_{k-1} -pq^\gamma \cdot a_{k-\gamma-1} &\quad\text{otherwise.}
\end{cases}
\end{equation}

These results are consistent: for $k < \gamma$, the read is too short so
the probability that it contains no seed is 1; for $k = \gamma$, the
read contains a seed if and only if it has no error, which occurs with
probability $q^\gamma$.

@@

As in the previous example, we need to know whether a nucleotide is a
sequencing error, but this time we also need to know its phase in the
repeated cycles of skipped positions. For this, we define the skip-$n$
alphabet $\mathcal{A}_n = \{\square, |, \Downarrow_0, \Downarrow_1,
\ldots, \Downarrow_n \}$. Again, the symbol $\square$ represents a correct
nucleotide and the symbol $|$ is a terminator added at the end of the
read. The symbols $\Downarrow_i$ ($i \in \mathbb{N}$) represent sequencing
errors and $i$ indicates the number of nucleotides till the next
non-skipped position (\textit{i.e.} $i=0$ for nucleotides immediately
before a non-skipped position and $i=n$ for nucleotides at a non-skipped
position).

@@

As mentioned above, after the terminator $\Downarrow_i$, we can append
segments with up to $\gamma+i-1$ symbols $\square$. The terminators of
those $\gamma+i$ possible segments are distributed according to the laws
of the arithmetic modulo $n+1$. When $i > 0$, for instance, if the next
segment has no symbol $\square$ then it is terminated by
$\Downarrow_{i-1}$. But the same goes if the segment has $n+1$ symbols
$\square$ (assuming $n+1 < \gamma$). Specifying the transfer matrix thus
involves a substantial amount of bookkeeping. One can check that the
expression of the transfer matrix $M_n(z)$ is
\begin{equation*}
\begin{blockarray}{cccccc}
   & \Dn{0} & \Dn{1} & \ldots & \Dn{n} & \nd \\
\begin{block}{c[ccccc]}
\Dn{0} & H_{0,0}(z) & H_{0,1}(z) & \ldots & H_{0,n}(z) & J_0(z) \\
\Dn{1} & H_{1,0}(z) & H_{1,1}(z) & \ldots & H_{1,n}(z) & J_1(z) \\
\vdots & \vdots & \vdots & \ddots & \vdots & \vdots \\
\Dn{n} & H_{n,0}(z) & H_{n,1}(z) & \ldots & H_{n,n}(z) & J_n(z) \\
\nd & 0 & 0 & \ldots & 0 & 0 \\
\end{block}
\end{blockarray},
\end{equation*}
\begin{eqnarray}
H_{i,j}(z) &=& pz \cdot (qz)^x \cdot \big( 1 + (qz)^{n+1} +
  \ldots + (qz)^{m(n+1)} \big), \\
\notag
  &\;& \text{where } x = i-j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma+i-1-x}{n+1} \right\rfloor, \\
J_i(z) &=& 1 + qz + (qz)^2 + \ldots + (qz)^{\gamma+i-1},
\end{eqnarray}
where $\lfloor \ldots \rfloor$ denotes the ``floor'' function.

@@

Since the duplicates are assumed to evolve independently of each other and
through the same mutagenesis process, the events are conditionally
independent (we do not have full independence because $S_0$ only depends
on sequencing errors whereas the events $S_j$ depend on sequencing errors
and on the divergence between duplicates). We can thus write
\begin{eqnarray}
\notag
P(\overline{S}_0 \cap \ldots \cap \overline{S}_N) &=&
  P(\overline{S}_0) \cdot P(\overline{S}_1 \cap \ldots \cap
  \overline{S}_N | \overline{S}_0) \\
\label{eq:Pnull}
  &=& P(\overline{S}_0) \cdot P(\overline{S}_1 | \overline{S}_0)^N =
  P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0 \cap
\overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{eqnarray}
Combinging the two equations above, we obtain
\begin{equation}
\label{eq:Poff}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{equation}

Hence, the probability that seeding is off target is a function of two
quantitites. We have already seen how to compute $P(\overline{S}_0)$ in
section~\ref{sec:example_exact} using recursive expression
(\ref{eq:recur}). We now need to find a way to compute $P(\overline{S}_0
\cap \overline{S}_1)$.

@@

\begin{definition}
At a given position of the read, a duplicate is a \emph{hard mask} if its
match length on the left side is strictly longer than the match length
of the target. A duplicate is a \emph{soft mask} if it has the same match
length as the target.
\end{definition}
